---
title: "Biostat 203B Homework 2"
subtitle: Due Feb 6 @ 11:59PM
author: Yi Fang
output: 
  html_document:
    toc: true
    toc_depth: 4 
---

Display machine information for reproducibility:
```{r}
sessionInfo()
```

```{r setup, message=F}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, cache.lazy = FALSE)
library(tidyverse)
library(data.table)
library(lubridate)
```

```{r}
os <- sessionInfo()$running
if (str_detect(os, "Linux")) {
  mimic_path <- "/mnt/mimiciv/1.0"
} else if (str_detect(os, "macOS")) {
  mimic_path <- "/Users/huazhou/Documents/Box Sync/MIMIC/mimic-iv-1.0"
}
```

In this exercise, we use tidyverse (ggpot2, dplyr, etc) to explore the [MIMIC-IV](https://mimic.mit.edu/docs/iv/) data introduced in [homework 1](https://ucla-biostat-203b.github.io/2022winter/hw/hw1/hw1.html) and to build a cohort of ICU stays.

```{r}
# tree -s -L 2 /Users/huazhou/Documents/Box\ Sync/MIMIC/mimic-iv-1.0
system(str_c("tree -s -L 2 ", shQuote(mimic_path)), intern = TRUE)
```

## Q1. `read.csv` (base R) vs `read_csv` (tidyverse) vs `fread` (data.table)

There are quite a few utilities in R for reading plain text data files. Let us test the speed of reading a moderate sized compressed csv file, `admissions.csv.gz`, by three programs: `read.csv` in base R, `read_csv` in tidyverse, and `fread` in the popular data.table package. 

Which function is fastest? Is there difference in the (default) parsed data types? (Hint: R function `system.time` measures run times.)

```{r}
admissions <- "/mnt/mimiciv/1.0/core/admissions.csv.gz"
#Base R
system.time(read.csv(admissions))
```
```{r}
admissions <- "/mnt/mimiciv/1.0/core/admissions.csv.gz"
#Tidyverse
system.time(read.csv(admissions))
```
```{r}
admissions <- "/mnt/mimiciv/1.0/core/admissions.csv.gz"
#fread
system.time(fread(admissions))
```
The base R use 40.8 seconds, tidyverse use 35.1 seconds, and 'fread' use 1.3 seconds. Fread is the fastest. I think the reason is that these commands work differently and 'fread' tries to first read rows into memory as character and then tries to convert them into integer and factor as data types.

For later questions, we stick to the tidyverse.

## Q2. ICU stays

`icustays.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/icustays/>) contains data about Intensive Care Units (ICU) stays. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/icu/icustays.csv.gz")), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Import `icustatys.csv.gz` as a tibble `icustays_tble`. 

```{r}
icustays <- read_csv("/mnt/mimiciv/1.0/icu/icustays.csv.gz", show_col_types = FALSE)
icustays_tble <- tibble(icustays)
```

2. How many unique `subject_id`? Can a `subject_id` have multiple ICU stays? 

```{r}
n_distinct(icustays_tble$subject_id)
head(count(icustays_tble, `subject_id`))
```
There are 53150 unique `subject_id` and a `subject_id` can have multiple ICU stays.

3. For each `subject_id`, let's only keep the first ICU stay in the tibble `icustays_tble`.

```{r}
icustays_tble <- distinct(icustays_tble, subject_id, .keep_all = TRUE)
```

## Q3. `admission` data

Information of the patients admitted into hospital is available in `admissions.csv.gz`. See <https://mimic.mit.edu/docs/iv/modules/core/admissions/> for details of each field in this file. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/core/admissions.csv.gz")), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Import `admissions.csv.gz` as a tibble `admissions_tble`.
```{r}
admissions_df <- read_csv(admissions, show_col_types = FALSE)
admissions_tble <- tibble(admissions_df)
```
2. Let's only keep the admissions that have a match in `icustays_tble` according to `subject_id` and `hadmi_id`.
```{r}
admissions_tble <- semi_join(admissions_tble, icustays_tble, by = c("subject_id", "hadm_id"))
```
3. Summarize the following variables by graphics. 

- admission year  
- admission month  
- admission month day  
- admission week day  
- admission hour (anything unusual?)  

```{r}
adyearplot <-
  ggplot(admissions_tble) +
  geom_density(aes(year(admittime))) +
  labs(x = "year", y = "frequency",
       title = "ICU Annually Admission Frequency Table") +
  theme(plot.title = element_text(hjust = 0.5))
adyearplot
```
```{r}
admonthplot <-
  ggplot(admissions_tble) +
  geom_bar(aes(month(admittime, label = TRUE)),
  fill = "skyblue",
  width = 0.6) +
  labs(x = "month", y = "frequency",
       title = "ICU Admission Frequency by Month") +
  theme(plot.title = element_text(hjust = 0.5))
admonthplot
```
```{r}
admdayplot <-
  ggplot(admissions_tble) +
  geom_bar(aes(mday(admittime)),
  fill = "skyblue",
  width = 0.6) +
  labs(x = "Day", y = "frequency",
       title = "ICU Admission Frequency by Day") +
  theme(plot.title = element_text(hjust = 0.5))
admdayplot
```
```{r}
adwkdayplot <-
  ggplot(admissions_tble) +
  geom_bar(aes(wday(admittime, label = TRUE, abbr = FALSE)),
  fill = "skyblue",
  width = 0.6) +
  labs(x = "Week Day", y = "frequency",
       title = "ICU Admission Frequency by Week Day") +
  theme(plot.title = element_text(hjust = 0.5))
adwkdayplot
```
```{r}
adhourplot <-
  ggplot(admissions_tble) +
  geom_bar(aes(hour(admittime)),
  fill = "skyblue",
  width = 0.6) +
  labs(x = "Hour", y = "frequency",
       title = "ICU Admission Frequency by Hour") +
  theme(plot.title = element_text(hjust = 0.5))
adhourplot
```
There is an unusual burst in ICU admission at 0 in the night, and I do not know why.

## Q4. `patients` data

Patient information is available in `patients.csv.gz`. See <https://mimic.mit.edu/docs/iv/modules/core/patients/> for details of each field in this file. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/core/patients.csv.gz")), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Import `patients.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/core/patients/>) as a tibble `patients_tble` and only keep the patients who have a match in `icustays_tble` (according to `subject_id`).
```{r}
patients_df <-  read_csv("/mnt/mimiciv/1.0/core/patients.csv.gz", show_col_types = FALSE)
patients_tble <- tibble(patients_df)
patients_tble <- semi_join(patients_tble, icustays_tble, by = "subject_id")
```
2. Summarize variables `gender` and `anchor_age`, and explain any patterns you see.
```{r}
genderageplot <-
  ggplot(patients_tble) +
  geom_density(aes(x = anchor_age)) +
  labs(x = "Age", y = "Frequency",
       title = "ICU Admission Age Frequency Table") +
  theme(plot.title = element_text(hjust = 0.5))
genderageplot
```
From this graph, I can tell that ICU has a high average age and the peak is around 67 years old. The chance of getting into ICU between 20 to 40 is low and are about the same. After 40 years old, it becomes more and more easier to get into ICU when the age increases.

```{r}
genderageplot <-
  ggplot(patients_tble) +
  geom_boxplot(aes(x = anchor_age, y = gender)) +
  labs(x = "Age", y = "Gender",
       title = "ICU Admission Gender and Age Box Plot") +
  theme(plot.title = element_text(hjust = 0.5))
genderageplot
```
From this graph, I can tell that female has a higher mean age of getting into ICU than male, and male gets into ICU earlier than female in their life on average. This is probably a side proof that female has a higher average lifetime than man do （because if you get into ICU later, it means that you are healthier.)

## Q5. Lab results

`labevents.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/hosp/labevents/>) contains all laboratory measurements for patients. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/hosp/labevents.csv.gz")), 
    " | head"
    ), 
  intern = TRUE
)
```
`d_labitems.csv.gz` is the dictionary of lab measurements. 
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/hosp/d_labitems.csv.gz")), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Find how many rows are in `labevents.csv.gz`.
```{r}
read_csv("/mnt/mimiciv/1.0/hosp/labevents.csv.gz",
           show_col_types = FALSE, col_select = c("labevent_id"), lazy = TRUE) %>%
  nrow()
```
There are 122103667 rows in `labevents.csv.gz`.

2. We are interested in the lab measurements of creatinine (50912), potassium (50971), sodium (50983), chloride (50902), bicarbonate (50882), hematocrit (51221), white blood cell count (51301), glucose (50931), magnesium (50960), and calcium (50893). Retrieve a subset of `labevents.csv.gz` only containing these items for the patients in `icustays_tble` as a tibble `labevents_tble`. 

    Hint: `labevents.csv.gz` is a data file too big to be read in by the `read_csv` function in its default setting. Utilize the `col_select` and `lazy` options in the `read_csv` function to reduce the memory burden.
```{r}
labevents_tble <- 
  read_csv("/mnt/mimiciv/1.0/hosp/labevents.csv.gz",
           show_col_types = FALSE, col_select = c("subject_id", "itemid"), lazy = TRUE) %>%
  filter(itemid == "50912" | itemid == "50917" | itemid == "50983" | itemid == "50902" 
         | itemid == "50882" | itemid == "51221" | itemid == "51301" 
         | itemid == "50931" | itemid == "50960" | itemid == "50893") %>%
  semi_join(icustays_tble, by = "subject_id")
```
```{r}
labevents_tble
```

3. Further restrict `labevents_tble` to the first lab measurement during the ICU stay. 
```{r}
labevents_tble <- distinct(labevents_tble, subject_id, .keep_all = TRUE)
```

4. Summarize the lab measurements by appropriate numerics and graphics.
```{r}

```

## Q6. Vitals from charted events

`chartevents.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/chartevents/>) contains all the charted data available for a patient. During their ICU stay, the primary repository of a patient’s information is their electronic chart. The `itemid` variable indicates a single measurement type in the database. The `value` variable is the value measured for `itemid`. The first 10 lines of `chartevents.csv.gz` are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/icu/chartevents.csv.gz")), 
    " | head"), 
  intern = TRUE
)
```
`d_items.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/d_items/>) is the dictionary for the `itemid` in `chartevents.csv.gz`. 
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/icu/d_items.csv.gz")), 
    " | head"), 
  intern = TRUE
)
```

1. We are interested in the vitals for ICU patients: heart rate (220045), mean non-invasive blood pressure (220181), systolic non-invasive blood pressure (220179), body temperature in Fahrenheit (223761), and respiratory rate (220210). Retrieve a subset of `chartevents.csv.gz` only containing these items for the patients in `icustays_tble` as a tibble `chartevents_tble`.

```{r}
chartevents_tble <- 
  read_csv("/mnt/mimiciv/1.0/icu/chartevents.csv.gz",
           show_col_types = FALSE, col_select = c("subject_id", "itemid"), lazy = TRUE) %>%
  filter(itemid == "220045" | itemid == "220181" | itemid == "220179" 
         | itemid == "223761" | itemid == "220210") %>%
  semi_join(icustays_tble, by = "subject_id")
```

2. Further restrict `chartevents_tble` to the first vital measurement during the ICU stay. 
```{r}
chartevents_tble <- distinct(chartevents_tble, subject_id, .keep_all = TRUE)
```
3. Summarize these vital measurements by appropriate numerics and graphics. 
```{r}

```

## Q7. Putting things together

Let us create a tibble `mimic_icu_cohort` for all ICU stays, where rows are  

- first ICU stay of each unique adult (age at admission > 18)

and columns contain at least following variables  

- all variables in `icustays.csv.gz`  
- all variables in `admission.csv.gz`  
- all variables in `patients.csv.gz`  
- first lab measurements during ICU stay  
- first vital measurements during ICU stay
- an indicator variable `thirty_day_mort` whether the patient died within 30 days of hospital admission (30 day mortality)

```{r}
mimic_icu_cohort <-
  inner_join(icustays_tble, admissions_tble, by = "subject_id") %>%
  inner_join(patients_tble, by = "subject_id") %>%
  inner_join(labevents_tble, by = "subject_id") %>%
  inner_join(chartevents_tble, by = "subject_id") %>%
  add_column(thirty_day_mort = if_else(difftime(deathtime, admittime, unit = "day") <= 30,
                                       "Died within 30 Days", "Not Died in 30 Days", "Not Died in 30 Days"))
```

## Q8. Exploratory data analysis (EDA)

Summarize following information using appropriate numerics or graphs.

- `thirty_day_mort` vs demographic variables (ethnicity, language, insurance, marital_status, gender, age at hospital admission)
```{r}

```

- `thirty_day_mort` vs first lab measurements
```{r}

```

- `thirty_day_mort` vs first vital measurements
```{r}

```

- `thirty_day_mort` vs first ICU unit
```{r}

```
